Peptide retention time prediction

### 2014-11-19 retention.tex
first stab on a discussion

\documentclass[a4paper]{article}
\usepackage[margin=0.5in]{geometry} 
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{multirow}

@@

\section{Introduction}
Mass spectrometry-based methods are unquestionably the most popular techniques for studying the proteint content of biological samples. The palette of protemic applications where such methods are used is getting increasingly diversified, ranging from identifying and quantifying
proteins in biological samples \cite{Geiger2012} to determining
posttranslational modifications \cite{Huttlin2010} and elucidating
protein interaction networks \cite{Gavin2011}.

\vspace{0.15cm}

@@

\vspace{0.15cm}

Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data. This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to interpret this data a major area of proteomics research. A particular difficult challenge in this context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
abundance~\cite{Angel2012}. Furthermore, a large proportion of
proteins are subjected to posttranslational modifications. As an
example, a few years ago more than 25,000 phosphorylation sites on
human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
 increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments. An example of such information is retention time, i.e. the time point when a peptide elutes from the liquid chromatography column.

\vspace{0.15cm}

@@

\vspace{0.15cm}

@@

\vspace{0.15cm}

Following these observations, a variety of retention time prediction
methods has been proposed. These methods can be roughly divided into
three classes: index-based methods, modeling-based techniques and
machine learning approaches. The main ideas underlying each category
are described below, while Table \ref{tab:rtmethods} lists a few of the most popular retention time 
prediction methods from each category.

@@

Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
\hline
 \end{tabular} \\
\end{table}

@@

\vspace{0.15cm}
% SSRcalc was proposed including a number of additional terms; then it was extended
% for a few other column
Following these observations, Krokhin et al. developed
SSRCalc \cite{Krokhin2004}, which is currently the most widely used
index-based retention time predictor. In the initial version of the
algorithm a collection of 346 tryptic peptides was used to derive two
sets of retention coefficients, one corresponding to the N-terminal
and one to all the other positions in a peptide sequence. Furthermore,
correction factors for peptide length and overall hydrophobicity were
included. Subsequent versions of the algorithm included terms
accounting for additional properties such as isoelectric point,
nearest-neighbor effects of charged peptides and propensity to form
helical structures \cite{Krokhin2006}. However, a limitation of the
approach is that the algorithm was optimized for predicting retention
times for certain chromatography setups. Thus, SSRCalc's prediction
ability deteriorates for experimental conditions diverging from the
setup on which the algorithm was optimized \cite{Spicer2007}.

@@

>>>>

\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{multirow}

@@

\setlength{\parskip}{0.15cm}

\section{Introduction}

Mass spectrometry-based methods are unquestionably the most popular
techniques for studying the proteint content of biological samples.
The palette of protemic applications where such methods are used is
getting increasingly diversified, ranging from identifying and
quantifying proteins in biological samples \cite{Geiger2012} to
determining posttranslational modifications \cite{Huttlin2010} and
elucidating protein interaction networks \cite{Gavin2011}.

@@

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications. As an example, a few years ago more
than 25,000 phosphorylation sites on human proteins were already
reported~\cite{Lemeer2009}. To manage this formidable diversity,
increasing efforts are made to extract as much information as possible
from an experiment.

@@

@@

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could

These methods can be roughly divided into four classes: index-based
methods, modeling-based techniques, machine learning approaches and
look-up tables-based approaches. The main ideas underlying each
category are described below, while Table \ref{tab:rtmethods} lists a
few of the most popular retention time prediction methods from each
category.

\subsection{Methods based on look-ups of previous registered retention times}

\cite{irt}

@@

Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
iRT-calculator & http://www.biognosys.ch/products/irt-calculator.html
& \cite{irt} \\
\hline
 \end{tabular} \\
\end{table}

@@

% SSRcalc was proposed including a number of additional terms; then it was extended
% for a few other column
Following these observations, Krokhin et al. developed
SSRCalc \cite{Krokhin2004}, which is currently the most widely used
index-based retention time predictor. In the initial version of the
algorithm a collection of 346 tryptic peptides was used to derive two
sets of retention coefficients, one corresponding to the N-terminal
and one to all the other positions in a peptide sequence. Furthermore,
correction factors for peptide length and overall hydrophobicity were
included. Subsequent versions of the algorithm included terms
accounting for additional properties such as isoelectric point,
nearest-neighbor effects of charged peptides and propensity to form
helical structures \cite{Krokhin2006}. However, a limitation of the
approach is that the algorithm was optimized for predicting retention
times for certain chromatography setups. Thus, SSRCalc's prediction
ability deteriorates for experimental conditions diverging from the
setup on which the algorithm was optimized \cite{Spicer2007}.

@@

\section{Discussion}

Interpretation of mass spectrometry data is a question of
understanding the properties that explain the behavior of peptides in
an mass spectrometry experiment. The available technologies to infer
peptides and proteins from mass spectrometry data involves predicting
properties of peptides in a database, and compare those properties to
the observed data. Such comparisons are for discovery proteomics done
by so called search engines, that traditionally are limited to
predicting predicting mass-to-charge rates of the precursor peptides,
and their fragmentation patterns. However, an attractive way to
increase the sensitivity of the inference procedure is to also predict
properties such as their iso-electric point~\cite{branca2014hirief} or
of particular interest for this review, chromatographic retention
time\cite{cerqueira2010mude}.

The principle of predictability also extends to targeted proteomics,
in SRM analysis the peptides retention time and fragments are
predicted from retention time, often derived from previous shotgun
analysis. As targeted proteomics move into more comprehensive
techniques, such as DIA/SWATH analysis, it becomes increasingly
attractive to use techniques that do not rely on previous observations
of the exact peptide to determine its properties, but instead to use
de-novo predictions from amino acid sequences.

% New technology to learn 
% - ranking-based method
% - deep learning
